---
title: "0. Creating a synthetic dataset"
author: "Charlotte Boys"
date: "2023-06-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

To demonstrate the inference of eGFR slopes with a joint modelling approach using open-source R packages, we first create a synthetic dataset. We simulate a patient cohort consisting of 100 patients sampled from two disease groups "Disease A" and "Disease B". 

The support functions which we use to simulate our data can be found in the script `code/syntheticData.R`.

Load necessary libraries:
```{r}
library(tidyverse)
```

Load support functions:
```{r}
source("code/syntheticData.R")
```

## Simulate the dataset

Simulate patient metadata for our synthetic cohort:
```{r}
patient_metadata <- sim_cohort_metadata(n_patients = 100)
```

```{r}
knitr::kable(head(patient_metadata))
```

Simulate longitudinal data:
```{r}
longitudinal_data <- sim_cohort_longitudinal_data(n_patients = 100,
                                                  patient_metadata = patient_metadata)
```

```{r}
knitr::kable(head(longitudinal_data))
```

The simulated longitudinal data currently contains two types of measurements:
```{r}
unique(longitudinal_data$type)
```

We have a range of follow-up times. Some dropout is random, due to time of biopsy relative to the end of follow-up, and we also have dropout not-at-random due to dialysis.

```{r}
endpoint_date <- longitudinal_data %>%
  dplyr::filter(type == "Dialysis") %>%
  dplyr::transmute(patient_id = patient_id, endpoint_years = years_from_biopsy)

to_plot <- longitudinal_data %>%
  dplyr::full_join(endpoint_date, by = "patient_id") %>%
  dplyr::group_by(patient_id) %>%
  dplyr::summarise(start = 0,
                   end = max(years_from_biopsy),
                   endpoint = min(endpoint_years))

to_plot %>%
  dplyr::mutate(patient_id = forcats::fct_reorder(patient_id, end, .desc = T)) %>% 
  ggplot() +
  geom_segment(aes(x=start,
                     xend=end,
                     y=patient_id,
                     yend=patient_id),
               linewidth=1, lineend = "round") +
  geom_point(aes(x = endpoint, 
                 y = patient_id, color = "Dialysis" ),
             na.rm=TRUE) + 
  labs(x = "Follow-up time", y = "Patient", color = "Endpoint reached") +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = c(0.85, 0.93))

```

Plot longitudinal SCr measurements for the first eight patients:
```{r}
longitudinal_data %>%
  dplyr::filter(patient_id %in% 1:8) %>%
  ggplot(aes(x = years_from_biopsy,
             y = measurement)) +
  geom_point(size = 1) +
  xlim(0, 20) +
  facet_wrap(vars(patient_id), nrow = 2)
```

Save patient metadata and longitudinal data:
```{r}
write.csv(patient_metadata, "data/simulated_metadata.csv", row.names = FALSE)
write.csv(longitudinal_data, "data/simulated_longitudinal_data.csv", row.names = FALSE)
```

