---
title: "02. Calculate eGFR"
author: "Charlotte Boys"
date: "2023-06-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this part of the tutorial, we calculate eGFR from serum creatinine measurements (umol/L) according to the CKD-EPI 2021 formula.

Load necessary libraries:
```{r}
library(tidyverse)
```

Load simulated longitudinal data and patient metadata:

```{r}
patient_metadata <- read.csv("data/simulated_metadata.csv")
longitudinal_data <- read.csv("data/simulated_longitudinal_data.csv")
```

Load support functions:

```{r}
source("code/calculateEGFR.R")
```

## Calculate eGFR

To calculate eGFR from creatinine serum measurements, we need information about:

- age at the time the measurement was taken
- sex

We take this from the patient metadata.

There are parameters within the equation - kappa, alpha, and a multiplication factor we'll call f - which are adjusted for sex. We add these to the dataframe, and use the CKD-EPI 2021 formula to calculate eGFR.
```{r}
serum_creatinine_egfr <- longitudinal_data %>% 
  dplyr::filter(type == "SCr") %>%
  dplyr::inner_join(patient_metadata,
                    by = "patient_id") %>%
  dplyr::mutate(kappa = ifelse(sex == "F", 61.894, 79.578),
                f = ifelse(sex == "F", 1.012, 1),
                alpha = ifelse(sex == "F", -0.241, -0.302),
                age = age_at_biopsy + years_from_biopsy) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(eGFR = CKD_EPI_2021(measurement, kappa, alpha, f, age)) %>%
  dplyr::ungroup()
```

Add eGFR measurements to the longitudinal data:

```{r}
longitudinal_egfr <- serum_creatinine_egfr %>%
  dplyr::transmute(patient_id = patient_id, 
                   type = "eGFR",
                   measurement = eGFR, 
                   units = "ml/min/1.73m²",
                   years_from_biopsy = years_from_biopsy,
                   days_from_biopsy = days_from_biopsy)

longitudinal_data <- rbind(longitudinal_data, longitudinal_egfr)
```

Plot the eGFR observations for the first twenty patients in our synthetic cohort:

```{r}
longitudinal_egfr %>%
  dplyr::filter(patient_id %in% 1:20) %>%
  ggplot() +
    geom_point(aes(x = years_from_biopsy, y = measurement), size = .5) +
  facet_wrap(~patient_id)
```

## Save data

Save the longitudinal data, now containing both serum creatinine and eGFR measurements, for future use:

```{r}
write.csv(longitudinal_data, "output/longitudinal_data.csv", row.names = FALSE)
```

