---
title: "03. Prepare data for joint modelling"
author: "Charlotte Boys, Elif Doğan Dar"
date: "2023-06-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Next, we prepare the data so it is ready for joint modelling. Since we are interested in the chronic eGFR trajectory, we filter out the episodes flagged as AKI and in-patient events.

Load necessary libraries:

```{r}
library(tidyverse)
```

Load data:

```{r}
patient_metadata <- read.csv("data/simulated_metadata.csv")
flagged_episodes <- read.csv("output/flagged_episodes.csv")
longitudinal_data <- read.csv("output/longitudinal_data.csv")
```

## Filter out measurements during AKI episodes

First, let's remind ourselves of how the flagged episodes data looked:
```{r}
knitr::kable(head(flagged_episodes))
```

There were three types of flags:
```{r}
unique(flagged_episodes$flag)
```

We define a function to filter out the flagged episodes:
```{r}
filter_flagged_episodes <- function(longitudinal_data,
                                    flagged_episodes,
                                    flags = c("AKI and in-patient", "AKI")){
  to_remove <- flagged_episodes %>%
  dplyr::filter(flag %in% flags)
  
  patients_without_flags <- longitudinal_data %>%
  dplyr::filter(!(patient_id %in% to_remove$patient_id)) %>%
  dplyr::distinct()
  
  patients_with_flags <- longitudinal_data %>%
  dplyr::filter(patient_id %in% to_remove$patient_id)
  
  check_observations <- dplyr::left_join(patients_with_flags,
                          to_remove,
                          by = "patient_id",
                          multiple = "all",
                          relationship = "many-to-many") %>%
  dplyr::mutate(flagged = ifelse(days_from_biopsy >= start & days_from_biopsy <= end, TRUE, FALSE))

  patients_with_flags_keep <- check_observations %>%
    dplyr::group_by(patient_id, days_from_biopsy) %>%
    dplyr::filter(!any(flagged == TRUE)) %>%
    dplyr::select(-c(start, end, n_obs, flag, flagged)) %>%
    dplyr::distinct()
  
  patients_with_flags_remove <- check_observations %>%
    dplyr::group_by(patient_id, days_from_biopsy) %>%
    dplyr::filter(any(flagged == TRUE)) %>%
    dplyr::select(-c(start, end, n_obs, flag, flagged)) %>%
    dplyr::distinct()
  
  stopifnot("Sum of rows to keep and rows to remove does not equal original number of rows" = nrow(patients_with_flags_keep) + nrow(patients_with_flags_remove) == nrow(patients_with_flags))
  
  measurements_filtered <- rbind(patients_without_flags, patients_with_flags_keep)
  return(measurements_filtered)
}
```

We filter the eGFR measurements, removing those taken during episodes of care flagged as meeting both the AKI and in-patient criteria:
```{r}
eGFR_filtered <- filter_flagged_episodes(longitudinal_data,
                                         flagged_episodes,
                                         flags = c("AKI and in-patient")) %>%
  dplyr::filter(type == "eGFR")
```

## Prepare eGFR data for joint modelling

In our simulated longitudinal data, we only have one clinical endpoint event of interest: dialysis.
We gather information about which patients reached the clinical endpoint, at which event time:

```{r}
endpoints <- c("Dialysis")

event_time <- longitudinal_data %>% 
  dplyr::filter(type %in% endpoints) %>%
  dplyr::transmute(patient_id, event_time = days_from_biopsy)

reached_endpoint <- event_time$patient_id
```

Add information about endpoints and final measurement time to the patient metadata:
```{r}
final_measurement <- longitudinal_data %>%
  dplyr::group_by(patient_id) %>%
  dplyr::slice_max(days_from_biopsy, n = 1, with_ties = F) %>% 
  dplyr::transmute(patient_id, last_measurement = days_from_biopsy)

patient_metadata <- patient_metadata %>% 
  dplyr::full_join(final_measurement, 
                   by = "patient_id" ,
                   multiple = "all") %>%
  dplyr::full_join(event_time, by = "patient_id", multiple = "all") %>%
  dplyr::distinct() %>%
  dplyr::group_by(patient_id) %>%
  dplyr::slice_head(n = 1) # Keep only one entry per patient_id
```

We add the patient metadata to the filtered longitudinal eGFR data, and also rescale measurements from days to years.
We add the columns `start`, `stop` and `endpoint` as they are needed for the joint modelling:

```{r}
eGFR_plus_metadata <- dplyr::full_join(eGFR_filtered, 
                                 patient_metadata, 
                                 by = c('patient_id')) %>%
  dplyr::mutate(last_measurement_y = last_measurement/365.25,
                event_time_y = event_time/365.25) %>%
  dplyr::distinct() %>%
  tidyr::drop_na(years_from_biopsy) %>%
  dplyr::group_by(patient_id) %>%
  dplyr::arrange(years_from_biopsy) %>%
  dplyr::mutate(start = years_from_biopsy) %>%
  dplyr::mutate(stop = dplyr::lead(start,
                                   order_by = years_from_biopsy,
                                   default = unique(last_measurement_y))) %>%
  dplyr::mutate(event = ifelse(is.na(event_time_y),
                               0,
                               ifelse(abs(stop-event_time_y) < 1e-10,
                                      1,
                                      0))) %>% 
  dplyr::mutate(endpoint = ifelse(patient_id %in% reached_endpoint,
                                  1, 
                                  0))
```

We have just a few more changes to make to be able to run the joint modelling.

`Surv()` requires the start time to be different from the stop time of the interval. However, when an observation is
the last available measurement for that patient, the start and the stop time are equal. As a work around, we omit the final observations from the model:

```{r}
to_omit <- eGFR_plus_metadata %>% dplyr::filter(stop <= start)
eGFR_plus_metadata <- setdiff(eGFR_plus_metadata, to_omit)
```

JM does not accept start times of zero. We have to set them to a small positive number.
Also, create a minimal dataframe of only the variables we will use in the modelling process, check that `patient_id` is numeric, and order the data according to `patient_id` and `start` time:

```{r}
eGFR_minimal <- eGFR_plus_metadata %>% 
  dplyr::select(patient_id, last_measurement_y, endpoint, measurement, disease, age_at_biopsy, sex, start, stop, event) %>%
  dplyr::mutate(start = ifelse(start < 1/365.25, 1e-5, start))

stopifnot("`patient_id` is not numeric" = all(is.numeric(eGFR_minimal$patient_id)))

# Order the longitudinal data according to the patient_id, and order the time points within each subject with respect to time
eGFR_minimal <- eGFR_minimal[order(eGFR_minimal$patient_id, eGFR_minimal$start),]
```

```{r}
knitr::kable(head(eGFR_minimal))
```

## Save data

```{r}
write.csv(eGFR_minimal, "output/eGFR_minimal.csv", row.names = FALSE)
```

