---
title: "5. Extract eGFR slopes"
author: "Charlotte Boys"
date: "2023-06-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Finally, we put together a report containing the eGFR measurements, flagged episodes and eGFR slopes for each patient.

Load necessary libraries:
```{r}
library(tidyverse)
```

Load support functions:
```{r}
source("code/extractSlopes.R")
```

Load data:
```{r}
pred_long <- read.csv("output/pred_long.csv")
longitudinal_data <- read.csv("output/longitudinal_data.csv")
flagged_episodes <- read.csv("output/flagged_episodes.csv")
```

## Plot eGFR report by patient

The support function `plot_by_patient()` allows us to generate a report plot showing the observed eGFR values, the fitted slope, and the flagged AKI episodes.

Prepare the data:
```{r}
predicted_values <- pred_long %>%
  dplyr::select(-c("age_at_biopsy", "last_measurement_y",
                   "low_measurement", "upp_measurement", "endpoint")) %>%
  dplyr::mutate(years_from_biopsy = start)

eGFR_values <- longitudinal_data %>%
  dplyr::filter(type == "eGFR") %>%
  dplyr::left_join(predicted_values, by = c("patient_id", "measurement", "years_from_biopsy"))

flagged_AKI <- flagged_episodes %>%
  dplyr::filter(start >= 0, flag == "AKI and in-patient") %>%
  dplyr::mutate(flag = "Meets AKI criteria\n& potential\nin-patient stay")
```

View plot for Patient 1:
```{r, warning = FALSE, message = FALSE}
plot_by_patient(patient_str = 1,
               longitudinal_data = eGFR_values,
               flagged_episodes = flagged_AKI)
```

Create a PDF report with all eGFR slopes:
```{r, warning = FALSE, message = FALSE}
all_patients <- unique(eGFR_values$patient_id)

plots <- lapply(all_patients,
                plot_by_patient,
                longitudinal_data = eGFR_values,
                flagged_episodes = flagged_AKI)

pdf("./output/follow_up_report_by_patient.pdf", width = 6, height = 4)
for (i in 1:length(plots)){
  plot(plots[[i]])
  }
dev.off()
```

## Extract table of eGFR slopes

Finally, the support function `extract_egfr_table()` allows us to extract a table of the eGFR slope (ml/min/1.73m²/year) and intercept (ml/min/1.73m²) according to the fitted joint mixed effects model.

```{r}
eGFR_slopes <- extract_egfr_table(patient_list = all_patients,
                         pred_eGFR = eGFR_values)
```

```{r}
knitr::kable(head(eGFR_slopes))
```

Save eGFR slopes:
```{r}
write.csv(eGFR_slopes, "output/fitted_eGFR_slopes.csv", row.names = FALSE)
```

